OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2021-10-29 paper/paper.Rmd

```{r deps, include=FALSE, message = FALSE}
library(assertthat)
library(cowplot)
library(glue)
library(here)
library(knitr)
library(schtools)
library(tidyverse)
set.seed(20210624)
snakemake <- params$snakemake
if (is_character(snakemake)) {
  stop("You need to knit this R Markdown document from the Snakemake workflow")
}
load(here("results", "stats.RData"))

@@

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering algorithm which uses the distance score
between all pairs of sequences in the dataset to cluster them into OTUs by
maximizing the Matthews Correlation Coefficient (MCC)
[@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_
methods such as the greedy clustering algorithms implemented in USEARCH and
VSEARCH, which only consider the distance between the query sequence and a
representative centroid sequence in the OTU [@edgar_search_2010;
@rognes_vsearch_2016].
In methods employing greedy clustering algorithms, only the distance between
each sequence and the centroid sequence is considered while clustering.
As a result, distances between pairs of sequences in the same OTU are frequenty
larger than the specified threshold, i.e. they are false positives.
In contrast, the OptiClust algorithm takes into account the distance between all
pairs of sequences when considering how to cluster sequences into OTUs and is
thus less willing to take on false positives.
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow and
memory requirements can be prohibitive for very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then clustered into OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

To overcome the limitations of current reference-based and _de novo_ clustering
algorithms while maintaining OTU quality, we developed OptiFit, a
reference-based clustering algorithm.
While other tools represent reference OTUs with a single sequence, OptiFit uses
multiple sequences in existing OTUs as the reference and fits new sequences
those reference OTUs.
In contrast to other tools, OptiFit considers all pairwise distance scores
between reference and query sequences when assigning sequences to OTUs in order
to produce OTUs of the highest possible quality.
Here, we tested the OptiFit algorithm with the reference as a public database
(e.g. Greengenes) or _de novo_ OTUs generated using a reference set from the
full dataset and compared the performance to existing tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
OptiFit is available within the mothur software program.

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton.
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from zero to one,
with a score of one occurring when all sequence pairs are true positives and
true negatives, and a score of zero when all pairs are false positives and false
negatives.
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are clustered into the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not clustered into the same OTU, and false positives if they are not in the same
OTU.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

@@

```{r db_names, include = FALSE}
db_names <- names(dn_dbs_mcc)
db_names_upper <- case_when(db_names == 'gg' ~ 'Greengenes',
                           TRUE ~ toupper(db_names))
close_enough(dn_dbs_mcc['gg'], dn_dbs_mcc['silva'])
close_enough(dn_dbs_mcc['gg'], dn_dbs_mcc['rdp'], 10^-1)
mcc_decreasing <- c(closed_fit_rdp_mcc, closed_fit_silva_mcc, closed_fit_gg_mcc,
                    open_fit_silva_mcc, open_fit_gg_mcc, open_fit_rdp_mcc,
                    opticlust_mcc)
assert_that(!is.unsorted(mcc_decreasing))
```

To test how OptiFit performs for reference-based clustering, we clustered each
dataset to three databases of reference OTUs: the Greengenes database, the SILVA
non-redundant database, and the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
Clustering sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and
`r closed_fit_silva_mcc` respectively, while the median MCC
was `r closed_fit_rdp_mcc` when clustering to RDP
(Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes,
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

```{r, include = FALSE}
assert_that(!is.unsorted(c(frac_fit_rdp, frac_fit_silva, frac_fit_gg)))
assert_that((frac_fit_gg - frac_fit_silva) < 10)
assert_that((frac_fit_gg - frac_fit_rdp) > 40)
```

@@

```{r, include = FALSE}
assert_that(!is.unsorted(c(closed_fit_silva_human_sec,
                           opticlust_human_sec,
                           open_fit_silva_human_sec)))
```

@@

```{r, include = FALSE}
assert_that(mcc_closed_fit_db_vs_vsearch > 10)
assert_that(frac_vsearch_vs_fit > 10)
assert_that((open_fit_gg_mcc - open_vsearch_mcc) > 0.20)
assert_that((sec_vsearch_vs_open_fit_db - sec_closed_fit_db_vs_vsearch) < 5)
```

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
<!--**TODO: delete or move this sentence:** _De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH in terms of MCC, but performed
`r sec_opticlust_vs_vsearch`% slower than VSEARCH.-->
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
reference database (Figure \ref{fig:results_sum}).
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of reference and query sequences in an OTU.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively.
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and
`r sec_vsearch_vs_open_fit_db`% on average, respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

```{r, include = FALSE}
close_enough(mcc_opticlust_vs_fit_split_simple, 0.02)
assert_that((frac_fit_split - frac_fit_gg) > 20)
assert_that(!is.unsorted(c(sec_closed_fit_split_vs_db, sec_closed_fit_db_vs_clust, sec_open_fit_split_vs_clust, sec_open_fit_split_vs_db)))
```

@@

```{r, include = FALSE}
close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
assert_that(!is.unsorted(c(sec_fit_split_dist,
                           sec_fit_split_simple,
                           sec_fit_split_abun)))
close_enough(frac_fit_split_simple, frac_fit_split_abun)
assert_that(frac_fit_split_dist < frac_fit_split_simple)
```

@@

```{r, include = FALSE}
ref_seqs <- read_tsv(here('subworkflows', '0_prep_db', 'data',
                          'seq_counts.tsv')) %>%
  pivot_wider(names_from = ref, values_from = nseqs) %>% as.list()
assert_that(ref_seqs[['rdp']] < ref_seqs[['silva']])
assert_that(ref_seqs[['silva']] < ref_seqs[['gg']])
```

>>>>

```{r deps, include=FALSE, message = FALSE}
library(assertthat)
library(cowplot)
library(glue)
library(here)
library(knitr)
library(schtools)
library(tidyverse)
set.seed(20210624)
snakemake <- params$snakemake
if (is_character(snakemake)) {
  stop("You need to knit this R Markdown document from the Snakemake workflow")
}

load(here("results", "stats.RData"))

@@

There are two main categories of OTU clustering algorithms: _de novo_ and
reference-based.
OptiClust is a _de novo_ clustering algorithm which uses the distance score
between all pairs of sequences in the dataset to cluster them into OTUs by
maximizing the Matthews Correlation Coefficient (MCC)
[@westcott_opticlust_2017].
This approach takes into account the distances between all pairs of sequences
when assigning query sequences to OTUs, in contrast to other _de novo_ methods
such as the greedy clustering algorithms implemented in USEARCH and VSEARCH
[@edgar_search_2010; @rognes_vsearch_2016].
In methods employing greedy clustering algorithms, only the distance between
each sequence and a representative centroid sequence in the OTU is considered
while clustering.
As a result, distances between pairs of sequences in the same OTU are frequently
larger than the specified threshold, i.e. they are false positives.
In contrast, the OptiClust algorithm takes into account the distance between all
pairs of sequences when considering how to cluster sequences into OTUs and is
thus less willing to take on false positives.
A limitation of _de novo_ clustering is that different OTU assignments will be
produced when new sequences are added to a dataset, making it difficult to use
_de novo_ clustering to compare OTUs between different studies.
Furthermore, since _de novo_ clustering requires calculating and comparing
distances between all sequences in a dataset, the execution time can be slow and
memory requirements can be prohibitive for very large datasets.
Reference clustering attempts to overcome the limitations of _de novo_
clustering methods by using a representative set of sequences from a database,
with each reference sequence seeding an OTU.
Commonly, the Greengenes set of representative full length sequences clustered
at 97% similarity is used as the reference with VSEARCH
[@desantis_greengenes_2006; @rognes_vsearch_2016; @noauthor_clustering_nodate].
Query sequences are then clustered into OTUs based on their similarity to the
reference sequences.
Any query sequences that are not within the distance threshold to any of the
reference sequences are either thrown out (closed reference clustering) or
clustered _de novo_ to create additional OTUs (open reference clustering).
While reference-based clustering is generally fast, it is limited by the
diversity of the reference database.
Novel sequences in the sample will be lost in closed reference mode if
they are not represented by a similar sequence in the database.
Previous studies found that the OptiClust _de novo_ clustering algorithm created
the highest quality OTU assignments of all clustering methods [@westcott_opticlust_2017].

To overcome the limitations of current reference-based and _de novo_ clustering
algorithms while maintaining OTU quality, we developed OptiFit, a
reference-based clustering algorithm.
While other tools represent reference OTUs with a single sequence, OptiFit uses
multiple sequences in existing OTUs as the reference and fits new sequences to 
those reference OTUs.
In contrast to other tools, OptiFit considers all pairwise distance scores
between reference and query sequences when assigning sequences to OTUs in order
to produce OTUs of the highest possible quality.
Here, we tested the OptiFit algorithm with the reference as a public database
(e.g. Greengenes) or _de novo_ OTUs generated using a reference set from the
full dataset and compared the performance to existing tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.
OptiFit is available within the mothur software program.

@@

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiClust first seeds each sequence into its own OTU as a singleton.
Then for each sequence, OptiClust considers whether the sequence should move to
a different OTU or remain in its current OTU, choosing the option that results
in a better Matthews correlation coefficient (MCC) [@westcott_opticlust_2017].
The MCC uses all values from a confusion matrix and ranges from negative one to
one, with a score of one occurring when all sequence pairs are true positives
and true negatives and a score of negative one occurring when all pairs are
false positives and false negatives.
Sequence pairs that are similar to each other (i.e. within the distance
threshold) are counted as true positives if they are clustered into the same OTU,
and false negatives if they are not in the the same OTU.
Sequence pairs that are not similar to each other are true negatives if they are
not clustered into the same OTU, and false positives if they are not in the same
OTU.
OptiClust iterations continue until the MCC stabilizes or until a maximum number
of iterations is reached.
This process produces _de novo_ OTU assignments with the most optimal MCC given
the input sequences.

@@

```{r assert_db_mccs, include = FALSE}
mcc_decreasing <- c(closed_fit_rdp_mcc, closed_fit_silva_mcc, closed_fit_gg_mcc,
                    open_fit_silva_mcc, open_fit_gg_mcc, open_fit_rdp_mcc,
                    opticlust_mcc)
assert_that(!is.unsorted(mcc_decreasing))
```

To test how OptiFit performs for reference-based clustering, we clustered each
dataset to three databases of reference OTUs: the Greengenes database, the SILVA
non-redundant database, and the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% using the V4 region of each
sequence (see Figure \ref{fig:workflow}).
After trimming to the V4 region, the databases contained 
`r dn_dbs %>% pull('nseqs')` unique sequences and produced _de novo_ MCC scores 
of `r dn_dbs %>% pull('med_mcc')` for `r dn_dbs %>% pull('refname')`, 
respectively.
Clustering sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and
`r closed_fit_silva_mcc` respectively, while the median MCC
was `r closed_fit_rdp_mcc` when clustering to RDP
(Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes,
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

@@

```{r assert_db_frac, include = FALSE}
assert_that(!is.unsorted(c(frac_fit_rdp, frac_fit_silva, frac_fit_gg)))
assert_that((frac_fit_gg - frac_fit_silva) < 10)
assert_that((frac_fit_gg - frac_fit_rdp) > 40)
```

@@

```{r assert_db_sec, include = FALSE}
assert_that(!is.unsorted(c(closed_fit_silva_human_sec,
                           opticlust_human_sec,
                           open_fit_silva_human_sec)))
```

@@

```{r assert_vsearch, include = FALSE}
assert_that(mcc_closed_fit_db_vs_vsearch > 10)
assert_that(frac_vsearch_vs_fit > 10)
assert_that((open_fit_gg_mcc - open_vsearch_mcc) > 0.20)
assert_that((sec_vsearch_vs_open_fit_db - sec_closed_fit_db_vs_vsearch) < 5)
```

To compare to the reference clustering methods used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
query sequence's similarity to the single reference sequence.
In contrast, OptiFit accepts reference OTUs which each may contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assigning sequences to OTUs.
In closed reference mode, OptiFit produced `r mcc_closed_fit_db_vs_vsearch`%
higher quality OTUs than VSEARCH, but VSEARCH was able to cluster
`r frac_vsearch_vs_fit`% more query sequences than OptiFit to the Greengenes
reference database (Figure \ref{fig:results_sum}).
This is because VSEARCH only considers the distances between each query sequence
to the single reference sequence, while OptiFit considers the distances between
all pairs of reference and query sequences in an OTU.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of
`r open_fit_gg_mcc` and `r open_vsearch_mcc`, respectively.
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and
`r sec_vsearch_vs_open_fit_db`% on average, respectively.
Thus, the more stringent OTU definition employed by OptiFit, which prefers the
query sequence to be similar to all other sequences in the OTU rather than to only one
sequence, resulted in fewer sequences being clustered to reference OTUs than when
using VSEARCH, but caused OptiFit to outperform VSEARCH in terms of both OTU
quality and execution time.

@@

```{r assert_split_simple, include = FALSE}
close_enough(mcc_opticlust_vs_fit_split_simple, 0.02)
assert_that((frac_fit_split - frac_fit_gg) > 20)
assert_that(!is.unsorted(c(sec_closed_fit_split_vs_db, sec_closed_fit_db_vs_clust, sec_open_fit_split_vs_clust, sec_open_fit_split_vs_db)))
```

@@

```{r assert_split_strat, include = FALSE}
close_enough(c(mcc_fit_split_simple, mcc_fit_split_abun), mcc_fit_split_dist)
assert_that(!is.unsorted(c(sec_fit_split_dist,
                           sec_fit_split_simple,
                           sec_fit_split_abun)))
close_enough(frac_fit_split_simple, frac_fit_split_abun)
assert_that(frac_fit_split_dist < frac_fit_split_simple)
```

@@

```{r assert_ref_sizes, include = FALSE}
ref_seqs <- dn_dbs %>% pull(nseqs) 
names(ref_seqs) <- dn_dbs %>% pull(ref)
assert_that(!is.unsorted(ref_seqs[['rdp']], 
                         ref_seqs[['silva']], 
                         ref_seqs[['gg']])
            )
```
